\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-15 doc/mlsys_paper/paper.tex
minor rewrites and suggestions for further improvements

As an example of usage, consider the linear regression objective
function\footnote{Just for simplicity, we ignore the bias term.  It can be
rederived by taking $x^*_i = (x_i, 1)$.}.  Given some dataset $X \in
\mathcal{R}^{n \times d}$ and associated responses $y \in \mathcal{R}^n$, the
model of linear regression is to assume that $y_i = x_i \theta$ for each
point and response $(x_i, y_i)$.  To fit this model $\theta \in \mathcal{R}^d$ best
to the data, we must find

\vspace*{-1.1em}
\begin{equation}
\operatorname{argmin}_\theta \sum_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-1.1em}

This objective function $f(\theta)$ has the associated gradient

\vspace*{-1.1em}
\begin{equation}
f'(\theta) = \sum_{i = 1}^n -2 x_i (y_i - x_i \theta) = -2 X^T (y - X \theta).
\end{equation}
\vspace*{-1.1em}

@@

% Indented a little to make it clear it's different from the text...
\vspace*{-0.4em}
\begin{minted}[fontsize=\small]{c++}
    class LinearRegressionFunction {
     public:
      // Construct the LinearRegressionFunction with the given data.
      LinearRegressionFunction(arma::mat& X, arma::vec& y) : X(X), y(y) {}

      // Compute the objective function.
      double Evaluate(const arma::mat& theta) {
        return arma::accu((y - X * theta) % (y - X * theta));
      }
      // Compute the gradient and store in 'gradient'.
      void Gradient(const arma::mat& theta, arma::mat& gradient) {
        gradient = -2 * X.t() * (y - X * theta);
      }
     private:
      arma::mat& X; arma::vec& y;
    };
\end{minted}
\vspace*{-0.4em}

This is all the implementation that is required to optimize this function with
any of {\tt ensmallen}'s gradient-based optimizers.  The code snippet below
shows how L-BFGS could be used to find the best parameters $\theta$:

\vspace*{-0.4em}
\begin{minted}[fontsize=\small]{c++}
    LinearRegressionFunction lrf(X, y); // we assume X and y are loaded elsewhere
    ens::L_BFGS lbfgs; // create L-BFGS optimizer with default parameters

@@

Only a slight variation on the signature of {\tt Evaluate()} and {\tt
Gradient()} would be needed to use the small-batch SGD-like optimizers that {\tt
ensmallen} provides.

>>>>

As an example of usage, consider the linear regression objective
function\footnote{For simplicity, we ignore the bias term.  It can be
rederived by taking $x^*_i = (x_i, 1)$.}.  Given a dataset $X \in
\mathcal{R}^{n \times d}$ and associated responses $y \in \mathcal{R}^n$, the
model of linear regression is to assume that $y_i = x_i \theta$ for each
point and response $(x_i, y_i)$.  To fit this model $\theta \in \mathcal{R}^d$
to the data, we must find

%% CS: i've added \nolimits to save a bit of space
\vspace*{-1.1em}
\begin{equation}
% f(\theta) =   %% CS: for clarity, perhaps we should have f(\theta) = ...
\operatorname{argmin}_\theta \sum\nolimits_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-1.1em}

The above objective function $f(\theta)$ has the associated gradient

\vspace*{-1.1em}
\begin{equation}
f'(\theta) = \sum\nolimits_{i = 1}^n -2 x_i (y_i - x_i \theta) = -2 X^T (y - X \theta).
\end{equation}
\vspace*{-1.1em}

@@

% Indented a little to make it clear it's different from the text...

%% CS: "accu((y - X * theta) % (y - X * theta))" is redoing computations
%% CS: ie. efficiency problem
%% CS: how about:
%% CS:   arma::mat tmp = y - X * theta;
%% CS:   return arma::accu(tmp % tmp);  // accumulate element-wise product
%% CS: in the above code, the comment hints at what the % operator does,
%% CS: as otherwise it's potentially confusing to someone not familiar with Armadillo
\vspace*{-0.4em}
\begin{minted}[fontsize=\small]{c++}
    class LinearRegressionFunction {
     public:
      // Construct the LinearRegressionFunction with the given data
      LinearRegressionFunction(arma::mat& X, arma::vec& y) : X(X), y(y) {}

      // Compute the objective function
      double Evaluate(const arma::mat& theta) {
        return arma::accu((y - X * theta) % (y - X * theta));
      }
      // Compute the gradient and store in 'gradient'
      void Gradient(const arma::mat& theta, arma::mat& gradient) {
        gradient = -2 * X.t() * (y - X * theta);
      }
     private:
      arma::mat& X; arma::vec& y;
    };
\end{minted}
\vspace*{-0.4em}
%% CS: the above code is currently shown across two pages - not a clean look.
%% CS: how about we place it in a figure ?

This is the entire required implementation to optimize the linear regression model with
any of {\tt ensmallen}'s gradient-based optimizers.  The code snippet below
shows how the L-BFGS optimizer can be used to find the best parameters $\theta$:

\vspace*{-0.4em}
\begin{minted}[fontsize=\small]{c++}
    LinearRegressionFunction lrf(X, y); // we assume X and y already hold data
    ens::L_BFGS lbfgs; // create L-BFGS optimizer with default parameters

@@

To use the small-batch SGD-like optimizers provided by {\tt ensmallen},
only a slight variation on the signature of {\tt Evaluate()} and {\tt
Gradient()} would be needed.
